---
title: "Tuesday"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("vegan")
library("reshape2")
library("outliers")
library("gdata") # for unmatrix
library("VIM") # For matrixplot
library("dplyr") 
```

Load the data we are using

```{r}
data("dune")
data("dune.env")
dummy_management <-
  as.data.frame(model.matrix(~ Management - 1, data = dune.env))
#add these to the dataset
dune.env.original <-
  dune.env #we keep a copy of the original version
dune.env <-
  dune.env %>% select(A1, Moisture, Manure, Use) %>% cbind(., dummy_management)
dune.env$Moisture <-
  as.numeric(as.character(dune.env$Moisture)) #make numeric
dune.env$Manure <- as.numeric(as.character(dune.env$Manure))
dune.env$Use <- as.numeric(dune.env$Use)
#make column names shorter
dune.env <-
  dune.env %>% rename(BF = ManagementBF,
                      HF = ManagementHF,
                      NM = ManagementNM,
                      SF = ManagementSF)
```

### [Question 7: Add one new species with cover degree 9 at the site with the lowest number of species. Do a new DCA-ordination.]{style="color:cornflowerblue"}

#### [Question:]{style="color:forestgreen"}

#### [7.1: What happens with the eigenvalues and ordination diagram?]{style="color:forestgreen"}

```{r ex6, inclue=TRUE, warning=FALSE, message=FALSE}
# calculate the number of species present in each plot
rowSums(dune != 0)
# find the plot with the lowest number of species
which.min(rowSums(dune != 0))
# create a new data frame with a new species with cover degree 9 at site 1
NewSp <- c(9, rep(0, 19))
SpeciesNewSp <- data.frame(dune, NewSp)
# run a new DCA with the extra species
DCANewSp <- decorana(SpeciesNewSp)
DCANewSp
```

```{r ex6plot, inclue=TRUE, warning=FALSE, message=FALSE}
# plot the DCA
ordiplot(DCANewSp, type="text", main="DCA with new sp.")
```

You may be surprised that you haven't got any total inertia values when printing decorana results, although in other software (e.g. CANOCO) these are available, together with percentage variance explained by particular axes. The reason for this is that DCA does not support the concept of total inertia values (also, it produces only four axes, i.e. four eigenvalues). See the comment in the code for the question 5.

------------------------------------------------------------------------

### [Question 8: Effects from increased variation.]{style="color:cornflowerblue"}

Add one new sample to the data, then add a cover of 9 for the existing species Chenopodium album "Chealbu" in the new sample. Make a new DCA-ordination. Next, remove the Chealbu = 9, and add instead a cover of 5 for the exisiting species Juncbufo and make a new DCA.

#### [Questions:]{style="color:forestgreen"}

#### [8.1 What happens with the eigenvalues and ordination after adding Chealbu and after adding Juncbufo?]{style="color:forestgreen"}

#### [8.2 Why did we choose to add these two species in the new sample?]{style="color:forestgreen"}

```{r ex7, inclue=TRUE, warning=FALSE, message=FALSE}
# figure out the column number for Chenopodium album (abbreviated "Chenalbu")
which(colnames(dune) == "Chenalbu")
# create a new data frame with a new sample containing Chenopodium album with a cover of 9
NewSampleChen <- c(rep(0, 7), 9, rep(0, 22))
SpeciesNewSampleChen <- data.frame(rbind(dune, NewSampleChen))
# run a new DCA with the extra sample
DCANewSampleChen <- decorana(SpeciesNewSampleChen)
DCANewSampleChen
# plot the DCA
ordiplot(DCANewSampleChen, type = "text", main = "DCA with new sample Chenopodium album")
# figure out the column number for Juncbufo
which(colnames(dune) == "Juncbufo")
# create a new data frame with a new sample containing Juncbufo with a cover of 5
NewSampleJunc <- c(rep(0, 15), 5, rep(0, 14))
SpeciesNewSampleJunc <- data.frame(rbind(dune, NewSampleJunc))
# run a new DCA with the extra sample
DCANewSampleJunc <- decorana(SpeciesNewSampleJunc)
DCANewSampleJunc
# plot the DCA
ordiplot(DCANewSampleJunc, type = "text", main = "DCA with new sample of Juncbufo")
```

------------------------------------------------------------------------

### [Question 9: Principal Coordinates Analysis (PCoA)]{style="color:cornflowerblue"}

Principal Coordinates Analysis (PCoA), also known as Multidimensional Scaling (MDS), is a generalization of the ordination techniques we have already seen as it can be estimated with any distance matrix. In contrast, PCA is restricted to Euclidean distances while CA preserves Chi-Square distances. Here, we will explore PCoA on four different distance matrices: (i) Bray-Curtis (ii) Jaccard (iii) Euclidean (iv) Mahalanobis

The R base statistics library offers the routine cmdscale() for performing MDS. There are numerous other libraries offering PCoA analysis

#### [Questions:]{style="color:forestgreen"}

#### [9.1 How similar are the ordination plots based on the four distance matrices?(remember question 2!)]{style="color:forestgreen"}

#### [9.2 Why are some plots more similar than others?]{style="color:forestgreen"}

#### [9.3 Which PCoA plot is most similar to the output from a PCA?]{style="color:forestgreen"}

#### [9.4 Which PCoA plot is most similar to the output from a CA?]{style="color:forestgreen"}

```{r}
# generate the four distance matrices
dune.bray.distance.matrix <- vegdist(dune, method = "bray")
dune.jaccard.distance.matrix <- vegdist(dune, method = "jaccard")
dune.euclidean.distance.matrix <- vegdist(dune, method = "euclidean")
dune.mahalanobis.distance.matrix <- vegdist(dune, method = "mahalanobis")
#
# what can be accomplished with the MDS routine?
#
help(cmdscale)
#
# run an MDS using cmdscale() for each of the four distance matrices
#
dune.bray.location <- cmdscale(dune.bray.distance.matrix)
dune.jaccard.location <- cmdscale(dune.jaccard.distance.matrix)
dune.euclidean.location <- cmdscale(dune.euclidean.distance.matrix)
dune.mahalanobis.location <-
  cmdscale(dune.mahalanobis.distance.matrix)
#
# now take a look at some plots
#
plot(dune.bray.location, xlab = "PCoA 1", ylab = "PCoA 2")
title(main = "PCoA on Bray Curtis distances of dune sites")
text(dune.bray.location,
     labels = row.names(dune.bray.location),
     pos = 2)
#
plot(dune.jaccard.location, xlab = "PCoA 1", ylab = "PCoA 2")
title(main = "PCoA on Jaccard distances of dune sites")
text(dune.jaccard.location,
     labels = row.names(dune.jaccard.location),
     pos = 2)
#
plot(dune.euclidean.location, xlab = "PCoA 1", ylab = "PCoA 2")
title(main = "PCoA on Euclidean distances of dune sites")
text(
  dune.euclidean.location,
  labels = row.names(dune.euclidean.location),
  pos = 2
)
#
plot(dune.mahalanobis.location, xlab = "PCoA 1", ylab = "PCoA 2")
title(main = "PCoA on Mahalanobis distances of dune sites")
text(
  dune.mahalanobis.location,
  labels = row.names(dune.mahalanobis.location),
  pos = 2
)
```

------------------------------------------------------------------------

### [Question 10: Non Metric Multidimensional Scaling]{style="color:cornflowerblue"}

NMDS does not use the absolute abundances of species in communities, but rather their *rank orders*. The use of ranks omits some of the issues associated with using absolute distance (e.g., sensitivity to transformation). At the conclusion of this exercise, you should be able to: State which metrics are supported in the vegan implementation of NMDS. Evaluate the appropriate number of dimensions for the ordination. Evaluate the consequences of using different distance metrics.

#### [Question:]{style="color:forestgreen"}

#### [10.1 Compare the vegan default NMDS to the results of a CA and a DCA.]{style="color:forestgreen"}

```{r ex8, inclue=TRUE, warning=FALSE, message=FALSE}
# run NMDS with the defaults and 2 dimensions, what is the stress?
#
dune.NMDS.bray.2 <- metaMDS(dune, k = 2)
#
# run NMDS with the defaults and 3 dimensions, what is the stress ?
#
dune.NMDS.bray.3 <- metaMDS(dune, k = 3)
#
# what happens when Euclidean distances are used?
#
dune.NMDS.euclidean.2 <- metaMDS(dune, distance = "euclidean", k = 2)
#
# plot some results
#
ordiplot(dune.NMDS.bray.2, type = "text", main = "NMDS from Bray Curtis distance, 2 Dimensions")
ordiplot(dune.NMDS.euclidean.2, type = "text", main = "NMDS from Euclidean distance, 2 Dimensions")
#
# create Shepard plots to show how the stress evolves
#
stressplot(dune.NMDS.bray.2)
stressplot(dune.NMDS.euclidean.2)
#
# generate the CA and DCA results
#
dune.CA <- cca(dune)
dune.DCA <- decorana(dune)
#
# plot the CA and DCA results
#
ordiplot(dune.CA, type = "text", main = "CA on Dune Meadow data")
ordiplot(dune.DCA, type = "text", main = "DCA on Dune Meadow data")


```

------------------------------------------------------------------------

### [Question 11: Data Quality Assessment]{style="color:cornflowerblue"}

These are only a few examples on how to assess data quality.

#### [Questions:]{style="color:forestgreen"}

#### [11.1 Is there a pattern in the distribution of missing values that may lead to the omission of a species or a sample?]{style="color:forestgreen"}

#### [11.2 Would you say that the original Dune Meadow data set is OK for further analyses? Motivate your answer!]{style="color:forestgreen"}

```{r ex10, inclue=TRUE, message=FALSE, warning=FALSE}
#data(dune)

# Create some outliers and missing values in a temporary dataset called mvo.dune (acronym for MultiVariate Outlier)
mvo.dune <- as.matrix(dune)
n_outl <- 5 # Give number of outliers to be created, here 5
v_outl <- 30 # Give value of the outlier(s), here 30
mvo.dune[sample(1:length(mvo.dune), n_outl, replace = FALSE)] <-
  v_outl # Replace some data with outliers, at random positions
mvo.dune[sample(1:length(mvo.dune), n_outl, replace = FALSE)] <-
  NA # Replace some data with NAs, at random positions

# Cleveland plot to inspect data
dotchart(as.matrix(mvo.dune), main = "Cleveland plot", xlab = "Observerd value")

# Find data > 9, as no value should be higher than 9 in the dune meadow data set
(GT9 <- which(mvo.dune > 9, arr.ind = TRUE))

# Find missing values
(mvo_miss <- which(is.na(mvo.dune), arr.ind = TRUE))

# Two ways of visualising missing data, using the package "VIM"
aggr(as.matrix(mvo.dune))
matrixplot(mvo.dune) 
```

We can also try testing for outliers in the data. Be aware that this is only indicative, and you still need to exercise your judgement about what should be excluded as an outlier!

```{r}

# Indicative search for outliers in a multivariate data set.
# Multiple peaks and high SD indicate outliers may occur
dist1 <-
  vegdist(
    mvo.dune,
    method = "euclidean",
    binary = FALSE,
    diag = FALSE,
    upper = FALSE,
    na.rm = TRUE
  )
d1 <- data.frame(unmatrix(as.matrix(dist1), byrow = TRUE))
d1r <- t(data.frame(strsplit(row.names(d1), ":")))
d1rr <- data.frame(d1, d1r)
d1mean <- tapply(d1rr$unmatrix.as.matrix.dist1.., d1rr$X1, mean)
sd(d1mean) # Standard deviation of the mean distances. If > 3: strong outlier(s)
plot(
  stats::density(d1mean),
  col = "red",
  xlab = "Distance between observations",
  main = "Indicative test of multivariate outliers\nData with noise"
)
legend("topright", paste("SD =", round(sd(d1mean), 1))) # Legend with standard deviation

# Same as above but on the original dune meadow dataset (i.e. without outliers)
dist1 <-
  vegdist(
    dune,
    method = "euclidean",
    binary = FALSE,
    diag = FALSE,
    upper = FALSE,
    na.rm = TRUE
  )
d1 <- data.frame(unmatrix(as.matrix(dist1), byrow = TRUE))
d1r <- t(data.frame(strsplit(row.names(d1), ":")))
d1rr <- data.frame(d1, d1r)
d1mean <- tapply(d1rr$unmatrix.as.matrix.dist1.., d1rr$X1, mean)
sd(d1mean) # Standard deviation of the mean distances. If > 3: strong outlier(s)
plot(
  stats::density(d1mean),
  col = "red",
  xlab = "Distance between observations",
  main = "Indictive test of multivariate outliers\nOriginal Dune Meadow data"
)
legend("topright", paste("SD =", round(sd(d1mean), 1))) # Legend with standard deviation

```

### Question 12: Messy ordination plots {style="color:cornflowerblue"}

By now you have seen a number of ordination plots, and probably realised that some plots get messy with a lot of species overlapping. To solve this, we can plot only those species that contribute the most to the ordination. The vegan function, goodness(), tells us how well species abundances are fitted by the ordination axes. We can then reduce the amount of information presented by labeling only those species which are well-fitted by the ordination axes being displayed. The function computes goodness of fit values for all species and you can then select a cut-off value for which species that should get a label (note that goodness of fit can also be useful information for purposes other than tidying up plots!). Here we try this with the CA from exercise 5 yesterday.

#### [Questions:]{style="color:forestgreen"}

#### [12.1 Is there any situation when this option should not be used?]{style="color:forestgreen"}

```{r}
dune.ca <- cca(dune)
# goodness of fit
gf <- goodness(dune.ca, model = "CA", display = "species", choices = 1:2)
head(gf)#the values in column PC2 are the cummulative goodness of fit values over the two axes
take <- gf[,2] >= 0.4 # which species to label. Here we choose species where fitted abundance on the two axes is at least 40%
scl <- "symmetric"    # choose scaling to use

# make an empty plot
plot(dune.ca, type = "n", scaling = scl, display = c("species", "sites"))

# add points for the species
points(dune.ca, scaling = scl, display = "species", cex = 0.7, col = "red", pch = "+")

# add points for the samples
points(dune.ca, scaling = scl, display = "sites", pch = NA)

# label the samples and species points
ordipointlabel(dune.ca, display = "sites", scaling = scl, add = TRUE)
ordipointlabel(dune.ca, display = "species", scaling = scl,col = "red", cex = 0.7, add = TRUE, select = take)

#compare with the original
plot(dune.ca)
```

------------------------------------------------------------------------
