---
title: "Friday"
editor_options: 
  markdown: 
    wrap: 72
---

Sometimes loaded packages in R can conflict with another package. Among
other issues, this can mean that the same function name (e.g. select())
is used in different packages, so the your code may give a different
output (or fail) depending on the packages you have loaded at the time!
By default, R uses the most recently loaded version of the function.
This can be over-ridden using the "conflicted" package. It is good
practice to write code that has a narrow purpose and loads the packages
needed for that purpose, rather than one huge script running many
different analyses. However, you may sometimes need to unload a package
or restart R to get the expected results. You can restart R (note: not
the same as restarting RStudio!) from the RStudio Session menu or using
the following code. The code is commented out because we don't actually
need to do this now, but it is good to know!

```{r}
#VIM package loaded on Tuesday conflicts with SIMPER function in Vegan which we will use today, if you wanted to use SIMPER after VIM you can detach VIM first
#detach("package:VIM", unload=TRUE, force = TRUE)

#Alternatively you can restart R and load only the packages you need
#.rs.restartR()
```

Now load the packages we need today

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(MASS)
library(tidyverse)
library(labdsv)
library(caret)
library(gllvm) #Generalised linear latent variable models
library(readxl)
library(mvabund)
library(lattice)
library(pls)

# pairwiseAdonis is not on CRAN, so we install it from GitHub. We use an updated
# fork of the original package (pmartinezarbizu/pairwiseAdonis), maintained for
# this course at https://github.com/JWeld/pairwiseAdonis
# The install only runs if the package is missing or older than the fork
# (version 0.4.2), so knitting does not reinstall it every time.
# remotes is a lightweight alternative to devtools for this purpose.
if (!requireNamespace("pairwiseAdonis", quietly = TRUE) ||
    packageVersion("pairwiseAdonis") < "0.4.2") {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("JWeld/pairwiseAdonis/pairwiseAdonis")
}
library(pairwiseAdonis)
```

Load the data we are using

```{r}
data("dune")
data("dune.env")
dummy_management <-
  as.data.frame(model.matrix(~ Management - 1, data = dune.env))
#add these to the dataset
dune.env.original <-
  dune.env #we keep a copy of the original version
dune.env <-
  dune.env %>% select(A1, Moisture, Manure, Use) %>% cbind(., dummy_management)
dune.env$Moisture <-
  as.numeric(as.character(dune.env$Moisture)) #make numeric
dune.env$Manure <- as.numeric(as.character(dune.env$Manure))
dune.env$Use <- as.numeric(dune.env$Use)
#make column names shorter
dune.env <-
  dune.env %>% rename(BF = ManagementBF,
                      HF = ManagementHF,
                      NM = ManagementNM,
                      SF = ManagementSF)
```

### [Exercise 29: PLS]{style="color:cornflowerblue"}

PLS regression, like PCA, seeks to find components which maximize the
variability of predictors but differs from PCA as PLS requires the
components to have maximum correlation with the response. The predictor
variables are mapped to a smaller set of variables, and within that
smaller space we perform a regression against the outcome variable. PLS
aims to choose new mapped variables that maximally explain the outcome
variable.

#### [Questions:]{style="color:forestgreen"}

#### [29.1 Which explanatory variables are most important to explain the response variables (A1 and Moisture)?]{style="color:forestgreen"}

#### [29.2 Are there any obvious outliers in the dataset?]{style="color:forestgreen"}

#### [29.3 Does the data seem to be in need of a transformation?]{style="color:forestgreen"}

Note that we use cross validation to find the number of components with
the lowest prediction error. For this small dataset that turns out to be
a single component, which makes for rather uninformative plots. In the
code below we therefore override the cross-validated choice and keep
three components for illustration. In a real analysis you would normally
stick with the cross-validated value, or at least justify a different
choice.

```{r PLS, echo=TRUE, message=FALSE, warning=FALSE}

#Note: loading pls masks the scores() function from vegan (the same kind of
#package conflict described at the start of today's notebook). If you go back to earlier exercises
#in the same session, call vegan::scores() explicitly.

#This function takes the same form as a lot of regression models in R:
#plsr(Response variable ~ Explanatory Variables, data = yourdata, scale = TRUE/FALSE)
#However the response variable can be a matrix of variables.

#Here we will use the dune species to predict environmental variables A1 and moisture.
#We have mostly done this in the other direction, but correlations go both ways! 
pls.response <- dplyr::select(dune.env.original, A1, Moisture)
#Moisture is stored as a factor with levels 1, 2, 4 and 5. as.numeric() on a
#factor returns the level codes (1, 2, 3, 4), not the values, so we go via
#as.character() to get the real moisture classes back.
pls.response$Moisture <- as.numeric(as.character(pls.response$Moisture))
pls.response <- as.matrix(pls.response)
pls.exp <- as.matrix(dune)
pls.fit <- plsr(pls.response ~ pls.exp,
                na.action = na.omit,
                validation = "LOO")
#Note that we have not scaled the responses. A1 (cm) has about 1.5 times the
#variance of Moisture (1-5 classes), so the PLS components are pulled somewhat
#more towards A1. To give the two responses equal weight, standardise them
#before fitting: pls.response <- scale(pls.response)
#(plsr's own scale = TRUE argument standardises the species, not the responses.)

summary(pls.fit)

#Cross validation is used to help us find the optimal number of retained dimensions.
#Then the model is rebuilt with this optimal number of dimensions.
cv <- RMSEP(pls.fit)

#cv$val is a 3-d array: estimate (CV / adjCV) x response (A1, Moisture) x
#number of components. Taking the adjCV slice gives a response x components matrix.
adjCV <- cv$val["adjCV", , ]
round(adjCV, 3)

#Each response on its own may choose a different number of components
#(here they happen to agree, but that is not guaranteed):
best.per.response <- apply(adjCV, 1, which.min) - 1
best.per.response

#To choose ONE number of components for the joint model, first express each
#response's error relative to its intercept-only error (so that A1, measured in
#cm, does not dominate Moisture, measured on a 1-5 scale), then average the two
#responses and take the minimum.
rel.adjCV <- adjCV / adjCV[, "(Intercept)"]
mean.rel <- colMeans(rel.adjCV)
round(mean.rel, 3)
best.dims <- unname(which.min(mean.rel) - 1)
best.dims

#Cross validation suggests a single component, which is a rather thin model
#for illustrating the plots below. For teaching purposes we keep three
#components instead. Try setting this to the cross-validated value and see
#how the results change!
best.dims <- 3

# Rerun the model with the chosen number of dimensions
pls.fit2 <-
  plsr(pls.response ~ pls.exp, ncomp = best.dims, na.action = na.omit, scale = TRUE)
summary(pls.fit2)

#Finally, we extract the useful information and format the output.
#coef() returns a species x response x ncomp array; take the species x response
#matrix for the fitted number of components (columns = A1 and Moisture).
coefficients <- coef(pls.fit2)[, , 1]

#Normalise EACH response separately so that the absolute values of the
#coefficients sum to 100 within a response. That makes the two responses
#comparable and means the bars really do show "percent of total effect".
coefficients <- sweep(coefficients, 2, colSums(abs(coefficients)), "/") * 100
colSums(abs(coefficients)) # check: both should be 100

#Plot the three strongest positive and three strongest negative predictors
#for each response (question 29.1 asks about both A1 and Moisture).
par(mfrow = c(2, 2), mar = c(6, 4, 3, 1))
for (resp in colnames(coefficients)) {
  co <- sort(coefficients[, resp])
  barplot(tail(co, 3), main = paste(resp, "- strongest positive"),
          las = 2, cex.names = 0.8)
  barplot(head(co, 3), main = paste(resp, "- strongest negative"),
          las = 2, cex.names = 0.8)
}
par(mfrow = c(1, 1))

#The next two plots use the full cross-validated fit (pls.fit) rather than
#pls.fit2, so that they always have three components to show whatever value
#you give best.dims above. PLS components are nested, so the first three
#components are identical in the two fits.

#This gives a pairwise plot of the correlation of each species with the three first components.
corrplot(pls.fit,
         comps = 1:3,
         labels = "names")

#This gives a pairwise plot of the score values for the three first components.
#Score plots are often used to look for patterns, groups or outliers in the data.
#The scores represent the different sites; labels = "names" prints the site numbers
#so that you can see which site is which (question 29.2).
plot(pls.fit, plottype = "scores", comps = 1:3, labels = "names")

#Study the predicted vs. measured plot to see if the data needs to be transformed.
plot(pls.fit2,
     ncomp = best.dims,
     asp = 1,
     line = TRUE)
```

### [Exercise 30: Differences between groups. ANOSIM, adonis and SIMPER]{style="color:cornflowerblue"}

ANOSIM (Analysis Of Similarities) is a non-parametric test of
significant difference between two or more groups, based on any distance
measure. In this case, we use management and change to Bray-Curtis
similarity index.

ANOSIM gives you the P value and an R value. R value close to 1
indicates high separation between levels of your factor while R value
close to 0 indicate no separation between levels of your factor.

Try also the recommended alternative "adonis" - Analysis of variance
using distance matrices - for partitioning distance matrices among
sources of variation and fitting linear models (e.g., factors,
polynomial regression) to distance matrices; uses a permutation test
with pseudo-F ratios.

#### [Questions:]{style="color:forestgreen"}

#### [30.1 What do the results from anosim and adonis tell you?]{style="color:forestgreen"}

#### [30.2 Give one other example of predefined groups that could be used as the grouping variable(s)!]{style="color:forestgreen"}

```{r}
# Start with a CA with Management as environmental factor and hulls around the different management types
dune.ca <- cca(dune)
plot(dune.ca, type = "n", scaling = "symmetric")
with(
  dune.env.original,
  points(
    dune.ca,
    display = "sites",
    scaling = "symmetric",
    col = as.numeric(Management),
    pch = 16
  )
)
with(
  dune.env.original,
  ordispider(dune.ca, Management, scaling = "symmetric", col = "skyblue")
)
with(dune.env.original,
     ordihull(dune.ca, Management, scaling = "symmetric", label = TRUE))

dune.dist <-
  vegdist(dune, method = "bray") #create distance matrix based on Bray-Curtis method
dune.ano <-
  anosim(dune.dist, dune.env.original$Management) #Comparing groupings based on management
summary(dune.ano)
plot(dune.ano, xlab = "Anosim, Dune meadow management types") #note the warning message here:
#notches are used in box plots to help visually assess whether the medians of distributions #differ. If the notches do not overlap, this is evidence that the medians are different.Here #the confidence region (the notch) went past the bounds (or hinges) of one of the boxes. 

#ANOSIM gives you the P value and a R value. R value close to 1 indicates high separation between levels of your factor while R value #close to 0 indicate no separation between levels of your factor.

```

Also try the recommended alternative "adonis" - Analysis of variance
using distance matrices - for partitioning distance matrices among
sources of variation and fitting linear models (e.g., factors,
polynomial regression) to distance matrices; this uses a permutation
test with pseudo-F ratios. We can also do pairwise comparisons to see
which groups differ from each other.

```{r}
dune.ado<-adonis2(formula = dune~Management, data = dune.env.original, method = "bray")
dune.ado

# You can also have more advanced models in adonis
dune.ado2<-adonis2(dune~Management*A1, by = "terms", data = dune.env.original)
dune.ado2

# adonis can also do the same kind of analysis as the CCA in exercise 21: does
# Management still explain anything once Manure is in the model?
# (by = "terms" tests the terms sequentially, so the order matters, as in CCA)
dune.ado3<-adonis2(dune~Manure+Management, by = "terms", data = dune.env.original)
dune.ado3

```

The adonis test tells us that there is a difference between our grouping
factors, but the pairwise test can tell us exactly which pairs of groups
differ from each other. Note that this only makes sense if the overall
test is significant!

#### [30.3 Which pairs of groups significantly differ from each other?]{style="color:forestgreen"}

```{r}
# Run the pairwiseAdonis function
pw <- pairwise.adonis2(dune~Management, data = dune.env.original)
pw
# Note that the R2 in each table is for that pair of groups only, so it is not
# comparable to the R2 of the overall adonis test above.

# With four groups we have run six tests, so we should correct the p-values for
# multiple testing, as we did after the forward selection on Wednesday.
# pairwise.adonis2() does not do this for us. Here we use the Holm correction.
pw.p <- sapply(pw[-1], function(x) x$`Pr(>F)`[1])
round(cbind(p = pw.p, p.holm = p.adjust(pw.p, method = "holm")), 3)
```

After a significant ANOSIM or PERMANOVA, you may want to know which
species are primarily responsible for the observed difference between
clusters. SIMPER (Similarity Percentage) will do this for you. The test
gives by-species p values for changes in abundance, but does not do
significance testing for overall difference between groups. In the
output tables, taxa are sorted in descending order of contribution to
group difference. Consider those contributions, not just the p values -
statistically significant does not mean ecologically important!
Especially if you have a large dataset, you may find some small
differences that are probably ecologically unimportant nevertheless
yield a p value under 0.05...

For a model-based test of which species differ between groups, see the mvabund
package that we use in exercise 34: manyglm() followed by anova(...,
p.uni = "adjusted") gives a test for each species.

#### [30.4 Compare the different groups and check which species contribute mostly to the difference between pairs of clusters.]{style="color:forestgreen"}

In the output there are columns for "ava" and "avb". As you may suspect,
these are average abundances of the species in group A and group B.

```{r message=FALSE, warning=FALSE}
sim <- simper(dune, dune.env.original$Management) #try management groupings
summary(sim)
```

Visually compare the result from simper with species positions in a
biplot.

```{r}
plot(dune.ca, type = "n", scaling = "species")
with(dune.env.original, points(dune.ca, display = "sites", scaling = "species", col = as.numeric(Management), pch=16))
with(dune.env.original, orditorp(dune.ca, display = "species", scaling = "species"))
with(dune.env.original, ordihull(dune.ca, Management, scaling = "species",label = TRUE))
```

A complementary/alternative approach to identifying which species are
most responsible for the differences between the groups is that of
indicator species - which species best characterise the groups or
clusters that we have identified?

In this exercise we will investigate if any particular species are
indicative for the four different management types.

#### [30.5 Use the resulting data table to find the two top indicator species for each of the four Management types!]{style="color:forestgreen"}

```{r, warning=FALSE, message=FALSE}
iva <- indval(dune, dune.env.original$Management) #this time we don't need dummy variables for management
iva.df <- as.data.frame(iva$indval)
#arrange in descending order to find best indicator species, change HF to other management types as needed
dplyr::arrange(rownames_to_column(iva.df), desc(HF))
```

------------------------------------------------------------------------

------------------------------------------------------------------------

### [Exercise 31: DFA]{style="color:cornflowerblue"}

The algorithm starts by finding directions that maximize the separation
between classes, then use these directions to predict the class of
individuals. These directions, called linear discriminants, are linear
combinations of predictor variables. The Dune dataset is too small to
really make use of this method (since we need to split the data into
test and training parts) so we will here use the classic "Iris" dataset
of different Iris plant species and the dimensions of different parts of
their flowers

#### [Questions:]{style="color:forestgreen"}

#### [31.1 How well does the algorithm distinguish between the species?]{style="color:forestgreen"}

#### [31.2 Here our groups are pre-defined (species). Would it be OK to use one of the clustering methods to find groups before running the DFA?]{style="color:forestgreen"}

```{r DFA}
# Load the data
data("iris")

# Split the data into training (80%) and test set (20%)
set.seed(123)
training.samples <- iris$Species %>%
  createDataPartition(p = 0.8, list = FALSE)
train.data <- iris[training.samples,]
test.data <- iris[-training.samples,]

# Discriminant analysis can be affected by the scale/unit in which predictor variables are measured. 
# It’s generally recommended to standardize/normalize continuous predictor before the analysis.
# Normalize the data. Categorical variables are automatically ignored.
# Estimate preprocessing parameters
preproc.param <- train.data %>%
  preProcess(method = c("center", "scale"))
# Transform the data using the estimated parameters
train.transformed <- preproc.param %>% predict(train.data)
test.transformed <- preproc.param %>% predict(test.data)

# Linear discriminant analysis - LDA ####
# Fit the model
model <- lda(Species ~ ., data = train.transformed)

model
plot(model, dimen = 1)#, type="both")
```

LDA determines group means and computes, for each individual, the
probability of belonging to the different groups. Each
observation/individual is then assigned to the group with the highest
probability score.

The lda() outputs contain the following elements:

Prior probabilities of groups: the proportion of training observations
in each group. For example 33% of the training observations are in the
Setosa group.

Group means: group center of gravity. Shows the mean of each variable in
each group.

Coefficients of linear discriminants: Shows the linear combination of
predictor variables that are used to form the LDA decision rule. If the
predictor variables are standardized before computing LDA, the
discriminator weights can be used as measures of variable importance for
feature selection.

Using the function plot() produces plots of the linear discriminants,
obtained by computing LD1 and LD2 for each of the training observations.

```{r}
# Make predictions
predictions <- model %>% predict(test.transformed)
# Model accuracy
mean(predictions$class==test.transformed$Species)

# The predict() function returns the following elements:
#   
# class: predicted classes of observations.
# posterior: is a matrix whose columns are the groups, rows are the individuals and values are the posterior probability that the corresponding observation belongs to the groups.
# x: contains the linear discriminants, described above

# Predicted classes
head(predictions$class, 6)
# Predicted probabilities of class membership.
head(predictions$posterior, 6) 
# Linear discriminants
head(predictions$x, 3) 

lda.data <- cbind(train.transformed, predict(model)$x)
#lda.data <- cbind(test.transformed, predictions$x)
ggplot(lda.data, aes(LD1, LD2)) +
  geom_point(aes(color = Species),size=3)+
  scale_colour_manual(name="",  
                      values = c("setosa"="blue", "versicolor"="orange", "virginica"="black"))

#More detailed assessment (probably more information than we need!)
#Calculates a cross-tabulation of predicted and observed classes with associated statistics.
confusionMatrix(predictions$class,test.data$Species)

```

------------------------------------------------------------------------

### [Exercise 32: Classification 2: Partition and Regression Trees]{style="color:cornflowerblue"}

Most of the code is taken from
<https://jonlefcheck.net/2015/02/06/a-practical-guide-to-machine-learning-in-ecology/>
Here, we will use the iris data. This is a "classic" data set used for
exploring multivariate methods for
discrimination.<https://en.wikipedia.org/wiki/Iris_flower_data_set>

#### [Questions:]{style="color:forestgreen"}

#### [32.1 Based on a visual analysis, which dimensions provide the best separation between iris species?]{style="color:forestgreen"}

#### [32.2 Are your insights from the visual analysis confirmed by the partition tree analysis?]{style="color:forestgreen"}

#### [32.3 Does the partition tree provide a perfect species identification or are there misclassifications?]{style="color:forestgreen"}

```{r}
# Before starting the analysis, we will generate pairwise scatterplots of the data
#
# the following line will first check to see inf the "psych" library is loaded, and only 
# load it if it is not present
#
if(!is.element("psych",installed.packages()[,1])) {install.packages("psych") }
library(psych)
library(rpart)
library(rpart.plot)
#
# create pairwise plots of the variables in the iris data set
#
pairs.panels(iris[,-5],gap=0,bg=c("red","blue","yellow")[iris$Species],pch=21)
#
# what does rpart() do ?
#
help(rpart)
#
# set up the partition tree model for the iris data
#
iris.tree.model<-rpart(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data = iris)
# 
# display the partition tree
#
rpart.plot(iris.tree.model, extra=0, box.palette =list("red", "blue", "forestgreen"))

#
# plots for confirmatory visual analysis
#
plot(iris$Petal.Length, col = c("red", "blue", "forestgreen")[iris$Species], xlab = "", ylab = "Petal Length (cm)")
legend(120, 2, c("setosa","versicolor","virginica"), col = c("red", "blue", "forestgreen"), pch = 1)
abline(h = 2.45)
#
plot(iris$Petal.Width, col = c("red", "blue", "forestgreen")[iris$Species], xlab = "", ylab = "Petal Width (cm)")
legend(20, 1.5, c("setosa","versicolor","virginica"), col = c("red", "blue", "forestgreen"), pch = 1)
abline(h = 1.75)
```

------------------------------------------------------------------------

### [Exercise 33: Model based ordination]{style="color:cornflowerblue"}

Dissimilarity-based ordination methods begin by calculating a
dissimilarity or distance matrix between sites (or more generally,
observational units). Afterwards, these are collapsed to a small number
of dimensions for plotting using an algorithm that attempts to preserve
and display information about these relative distances. The model-based
approach explicitly specifies a statistical model of the process that
generated the observed data. Model-based ordination methods (mostly) use
an extension of generalised linear mixed models known as latent variable
models. As the name suggests, these models account for the correlation
between taxa via the inclusion of a small number of latent variables.
Latent variables act as natural ordination axes reflecting unobserved
covariates.

#### [Question:]{style="color:forestgreen"}

#### [33.1 How do the ordinations produced compare to a distance based ordination (for example an NMDS) of the same species data?]{style="color:forestgreen"}

```{r}
#load species abundance data
spe<-as.data.frame(read_xlsx("Data/dune.xlsx", na = ".",sheet = 1, col_names = TRUE)); spe[1]<-NULL
#we need to provide the model with an appropriate probability distribution for the data.
#Given our abundance data is on a scale of 1 - 9 we could classify it as ordinal data,
#but we can try another distribution as well to compare results.

fitnb <- gllvm(y = spe, family = "negative.binomial") #This takes a minute or so to run!
fitord <- gllvm(y = spe, family = "ordinal", zeta.struc = "common")
#ignore the warning messages generated for now!

#As with "normal" linear regressions we can check model fit with widely used tools like
#Akaike Information Criterion (AIC). AIC aims to select the model which best explains 
#the variance in the dependent variable with the fewest number of independent variables
#So it helps select a simpler model over a complex model, if the complexity is not helpful.
#A lower AIC is better!

AIC(fitnb)
AIC(fitord)

#It seems the ordinal model is the best.
```

We can also use graphical checks of model fit, as with any regression.
If you are familiar with these try running plot(fitord) and plot(fitnb).

Now we can look at the ordination plot for the best model.

```{r}
par(mfrow = c(1, 2))
ordiplot.gllvm(
  fitord,
  main = NULL,
  biplot = T,
  s.colors = "black",
  spp.colors = "blue",
  alpha = 0.45,
  cex.spp = 0.7,
  jitter = F,
  s.cex = 0.6
)

abline(h = 0, v = 0, lty = 2)
#compare with an NMDS
distord <- metaMDS(spe)
vegan::ordiplot(distord, type = "t")
par(mfrow = c(1, 1))
```

We can do a lot more with model based ordinations, including
incorporating environmental variables. We can also include information
on traits and perform a fourth corner analysis.

------------------------------------------------------------------------

### [Exercise 34: Fourth Corner]{style="color:cornflowerblue"}

Here we will use another model based method (the "mvabund" package for
analysing multivariate abundance data) to perform a fourth corner
analysis. That is to say we will include trait information and
environmental variables, and investigate the relationship between plant
traits and environmental variables.

#### [Question:]{style="color:forestgreen"}

#### [34.1 Do the associations in the plot make sense ecologically?]{style="color:forestgreen"}

```{r}
# Read and treat data

env <- as.data.frame(read_xlsx("Data/dune_env.xlsx", na = "NA", sheet = 1, col_names = TRUE)); env[1] <- NULL
env$Management <- as.factor(env$Management)
env$Use <- as.factor(env$Use)

# This is from a database of plant traits
traits <- as.data.frame(read_xlsx("Data/TylerTraits.xlsx", na = "NA", sheet = 1, col_names = TRUE, skip = 1))
rownames(traits) <- traits$Spe; traits$Spe <- NULL
names(traits)

# Remove traits not feasible for this analysis. Select by name rather than by
# column position, so the code still does the right thing if the trait file
# gains or loses columns. If you want something back in the model, just delete it
# from this list.
drop_traits <- c("Biodiversity_relevance", "Nectar", "Heat", "Cold", "Temp",
                 "Continentality", "N_fixation")
missing_drop <- setdiff(drop_traits, names(traits))
if (length(missing_drop) > 0) {
  warning("Not found in trait file: ", paste(missing_drop, collapse = ", "))
}
traits <- traits[, setdiff(names(traits), drop_traits), drop = FALSE]

# traitglm() matches species to traits by position, not by name, so the rows of
# traits must be in exactly the same order as the columns of spe. Check that the
# two files contain the same species, then reorder traits to match spe.
stopifnot(setequal(colnames(spe), rownames(traits)))
traits <- traits[colnames(spe), , drop = FALSE]
stopifnot(identical(rownames(traits), colnames(spe)))

names(traits)

```

**Fourth corner analysis**

Given that there are many traits and even environmental factors that are
not strong predictors of abundance, we can reduce the number of
predictor terms in our models. LASSO is an optimization algorithm (Least
Absolute Shrinkage and Selection Operator) that is used in this case.

```{r}

fitfourth = traitglm(spe, env, traits, method = "glm1path")
fitfourth$fourth.corner #examine fourth corner terms, you may want to convert to a dataframe for further analysis
#View(as.data.frame(fitfourth$fourth.corner))

#and plot the results
a        = max(abs(fitfourth$fourth.corner))
colort   = colorRampPalette(c("blue", "white", "red"))
plot.4th = levelplot(
  t(as.matrix(fitfourth$fourth.corner)),
  xlab = "Environmental Variables",
  ylab = "Species traits",
  col.regions = colort(100),
  at = seq(-a, a, length = 100),
  scales = list(x = list(cex = 0.6, rot = 45), y = list(cex = 0.6))
)
print(plot.4th)

#Goodness of fit check - do the points lie mostly along the diagonal line? If not you may have a problem!
qqnorm(residuals(fitfourth)); abline(c(0,1),col="red") # for a normal quantile plot.


```

We can also test for significance and explore the effects of individual
trait by environment interactions using the summary() function of the
mvabund package. Warning! This code is commented out as it is very slow
to run with this many traits and environmental variables, even with a
limited number of iterations. Don't run this code chunk unless you want
to take a rather long coffee break! If this is something you do want to
do, narrow down the focus to fewer traits and environmental factors,
using only those you are really interested in.

```{r}
#We can test for overall significance using an anova (only on default manyglm model, 
#not available for Lasso model). This is a bit slow, but only several minutes.
#fitfourth2 = traitglm(spe, env, traits)
#pvalues <- anova.traitglm(fitfourth2,nBoot=20) #overall significant

#INDIVIDUAL TRAITS#
#We will limit the number of bootstrap reps to 5 (nBoot=5) but at least several hundred should be used for reliable results/final analyses. Even with nBoot=5 this takes a few minutes, so several hundred would take hours

#summary(fitfourth2, resamp = "montecarlo", nBoot=5)

#Example of output if you don't want to wait for the code to run...

#UsePasture.Grazing                        0.063     0.500

#The numbers returned are a test statistic and a P-value calculated by resampling rows of the data. In this case the interaction is not significant (p=0.5). Note that with only 5 reps the smallest possible P-value is 1/6 = 0.167, so nothing can be significant here - another reason to use many more reps in a real analysis
```

------------------------------------------------------------------------
