---
title: "Friday"
editor_options: 
  markdown: 
    wrap: 72
---

Sometimes loaded packages in R can conflict with another package. Among
other issues, this can mean that the same function name (e.g. select())
is used in different packages, so the your code may give a different
output (or fail) depending on the packages you have loaded at the time!
It is good practice to write code that has a narrow purpose and loads
the packages needed for that purpose, rather than one huge script
running many different analyses. However, you may sometimes need to
unload a package or restart R to get the expected results. You can
restart R (note: not the same as restarting RStudio!) from the RStudio
Session menu or using the following code. The code is commented out
because we don't actually need to do this now, but it is good to know!

```{r}
#VIM package loaded on Tuesday conflicts with SIMPER function in Vegan which we will use today, if you wanted to use SIMPER after VIM you can detach VIM first
#detach("package:VIM", unload=TRUE, force = TRUE)

#Alternatively you can restart R and load only the packages you need
#.rs.restartR()
```

Now load the packages we need today

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(MASS)
library(tidyverse)
library(labdsv)
library(caret)
library(gllvm) #Generalised linear latent variable models
library(readxl)
library(mvabund)
library(lattice)
library(pls)
```

Load the data we are using

```{r}
data("dune")
data("dune.env")
dummy_management <-
  as.data.frame(model.matrix(~ Management - 1, data = dune.env))
#add these to the dataset
dune.env.original <-
  dune.env #we keep a copy of the original version
dune.env <-
  dune.env %>% select(A1, Moisture, Manure, Use) %>% cbind(., dummy_management)
dune.env$Moisture <-
  as.numeric(as.character(dune.env$Moisture)) #make numeric
dune.env$Manure <- as.numeric(as.character(dune.env$Manure))
dune.env$Use <- as.numeric(dune.env$Use)
#make column names shorter
dune.env <-
  dune.env %>% rename(BF = ManagementBF,
                      HF = ManagementHF,
                      NM = ManagementNM,
                      SF = ManagementSF)
```

### [Exercise 30: Differences between groups. ANOSIM, adonis and SIMPER]{style="color:cornflowerblue"}

ANOSIM (Analysis Of Similarities) is a non-parametric test of
significant difference between two or more groups, based on any distance
measure. In this case, use the clusters from the K-means exercise.
Change to Bray-Curtis similarity index.

ANOSIM gives you the P value and an R value. R value close to 1
indicates high separation between levels of your factor while R value
close to 0 indicate no separation between levels of your factor.

Try also the recommended alternative "adonis" - Analysis of variance
using distance matrices - for partitioning distance matrices among
sources of variation and fitting linear models (e.g., factors,
polynomial regression) to distance matrices; uses a permutation test
with pseudo-F ratios.

#### [Questions:]{style="color:forestgreen"}

#### [30.1 What do the results from anosim and adonis tell you?]{style="color:forestgreen"}

#### [30.2 Give one other example of predefined groups that could be used as the grouping variable(s)!]{style="color:forestgreen"}

```{r}
# Start with a CA with Management as environmental factor and hulls around the different management types
dune.ca <- cca(dune)
fit.mgm <- envfit(dune.ca ~ Management, dune.env.original, perm = 0)
plot(dune.ca, type = "n", scaling = "symmetric")
with(
  dune.env.original,
  points(
    dune.ca,
    display = "sites",
    scaling = "symmetric",
    col = as.numeric(Management),
    pch = 16
  )
)
with(
  dune.env.original,
  ordispider(dune.ca, Management, scaling = "symmetric", col = "skyblue")
)
with(dune.env.original,
     ordihull(dune.ca, Management, scaling = "symmetric", label = TRUE))

dune.dist <-
  vegdist(dune, method = "bray") #create distance matrix based on Bray-Curtis method
dune.ano <-
  anosim(dune.dist, dune.env.original$Management) #Comparing groupings based on management
summary(dune.ano)
plot(dune.ano, xlab = "Anosim, Dune meadow management types") #note the error message here:
#notches are used in box plots to help visually assess whether the medians of distributions #differ. If the notches do not overlap, this is evidence that the medians are different.Here #the confidence region (the notch) went past the bounds (or hinges) of one of the boxes. 

#ANOSIM gives you the P value and a R value. R value close to 1 indicates high separation between levels of your factor while R value #close to 0 indicate no separation between levels of your factor.

# Try also the recommended alternative "adonis" - Analysis of variance using distance matrices - for partitioning distance matrices among sources of variation and fitting linear models (e.g., factors, polynomial regression) to distance matrices; uses a permutation test with pseudo-F ratios.

dune.ado<-adonis2(formula = dune~Management, data = dune.env.original, method = "bray")
dune.ado

# You can also have more advanced models in adonis
dune.ado2<-adonis2(dune~Management*A1, data = dune.env.original)
dune.ado2
```

After a significant ANOSIM, you may want to know which species are
primarily responsible for the observed difference between clusters.
SIMPER (Similarity Percentage) will do this for you. The test gives
by-species p values for changes in abundance, but does not do
significance testing for overall difference between groups. In the
output tables, taxa are sorted in descending order of contribution to
group difference. Consider those contributions (effect sizes), not just
the p values - statistically significant does not mean ecologically
important! Especially if you have a large dataset, you may find some
small differences that are probably ecologically unimportant
nevertheless yield a p value under 0.05...

#### [30.3 Compare the different groups and check which species contribute mostly to the difference between pairs of clusters.]{style="color:forestgreen"}

```{r message=FALSE, warning=FALSE}
sim <- simper(dune, dune.env.original$Management) #try management groupings
summary(sim)
```

Visually compare the result from simper with species positions in a
biplot.

```{r}
plot(dune.ca, type = "n", scaling = "species")
with(dune.env.original, points(dune.ca, display = "sites", scaling = "species", col = as.numeric(Management), pch=16))
with(dune.env.original, orditorp(dune.ca, display = "species", scaling = "species"))
with(dune.env.original, ordihull(dune.ca, Management, scaling = "species",label = TRUE))
```

A complementary/alternative approach to identifying which species are
most reposnisble for the differences between the groups is that of
indicator species - which species best characterise the groups or
clusters that we have identified?

In this exercise we will investigate if any particular species are
indicative for the four different management types.

#### [30.4 Use the resulting data table to find the two top indicator species for each of the four Management types!]{style="color:forestgreen"}

```{r, warning=FALSE, message=FALSE}
iva <- indval(dune, dune.env.original$Management) #this time we don't need dummy variables for management
iva.df <- as.data.frame(iva$indval)
#arrange in descending order to find best indicator species, change BF to other management types as needed
dplyr::arrange(rownames_to_column(iva.df), desc(HF))
```

------------------------------------------------------------------------

------------------------------------------------------------------------

### [Exercise 31: DFA]{style="color:cornflowerblue"}

The algorithm starts by finding directions that maximize the separation
between classes, then use these directions to predict the class of
individuals. These directions, called linear discriminants, are linear
combinations of predictor variables. The Dune dataset is too small to
really make use of this method (since we need to split the data into
test and training parts) so we will here use the classic "Iris" dataset
of different Iris plant species and the dimensions of different parts of
their flowers

#### [Questions:]{style="color:forestgreen"}

#### [31.1 How well does the algorithm distinguish between the species?]{style="color:forestgreen"}

#### [31.2 Here our groups are pre-defined (species). Would it be OK to use one of the clustering methods to find groups before running the DFA?]{style="color:forestgreen"}

```{r DFA}
# Load the data
data("iris")

# Split the data into training (80%) and test set (20%)
set.seed(123)
training.samples <- iris$Species %>%
  createDataPartition(p = 0.8, list = FALSE)
train.data <- iris[training.samples,]
test.data <- iris[-training.samples,]

# Discriminant analysis can be affected by the scale/unit in which predictor variables are measured. 
# It’s generally recommended to standardize/normalize continuous predictor before the analysis.
# Normalize the data. Categorical variables are automatically ignored.
# Estimate preprocessing parameters
preproc.param <- train.data %>%
  preProcess(method = c("center", "scale"))
# Transform the data using the estimated parameters
train.transformed <- preproc.param %>% predict(train.data)
test.transformed <- preproc.param %>% predict(test.data)

# Linear discriminant analysis - LDA ####
# Fit the model
model <- lda(Species ~ ., data = train.transformed)

model
plot(model, dimen = 1)#, type="both")
```

LDA determines group means and computes, for each individual, the
probability of belonging to the different groups. Each
observation/individual is then assigned to the group with the highest
probability score.

The lda() outputs contain the following elements:

Prior probabilities of groups: the proportion of training observations
in each group. For example 33% of the training observations are in the
Setosa group.

Group means: group center of gravity. Shows the mean of each variable in
each group.

Coefficients of linear discriminants: Shows the linear combination of
predictor variables that are used to form the LDA decision rule. If the
predictor variables are standardized before computing LDA, the
discriminator weights can be used as measures of variable importance for
feature selection.

Using the function plot() produces plots of the linear discriminants,
obtained by computing LD1 and LD2 for each of the training observations.

```{r}
# Make predictions
predictions <- model %>% predict(test.transformed)
# Model accuracy
mean(predictions$class==test.transformed$Species)

# The predict() function returns the following elements:
#   
# class: predicted classes of observations.
# posterior: is a matrix whose columns are the groups, rows are the individuals and values are the posterior probability that the corresponding observation belongs to the groups.
# x: contains the linear discriminants, described above

# Predicted classes
head(predictions$class, 6)
# Predicted probabilities of class membership.
head(predictions$posterior, 6) 
# Linear discriminants
head(predictions$x, 3) 

lda.data <- cbind(train.transformed, predict(model)$x)
#lda.data <- cbind(test.transformed, predictions$x)
ggplot(lda.data, aes(LD1, LD2)) +
  geom_point(aes(color = Species),size=3)+
  scale_colour_manual(name="",  
                      values = c("setosa"="blue", "versicolor"="orange", "virginica"="black"))

# Model accuracy
mean(predictions$class==test.transformed$Species)

#More detailed assessment (probably more information than we need!)
#Calculates a cross-tabulation of observed and predicted classes with associated statistics.
confusionMatrix(test.data$Species, predictions$class)

```

------------------------------------------------------------------------

### [Exercise 32: Model based ordination]{style="color:cornflowerblue"}

Dissimilarity-based ordination methods begin by calculating a
dissimilarity or distance matrix between sites (or more generally,
observational units). Afterwards, these are collapsed to a small number
of dimensions for plotting using an algorithm that attempts to preserve
and display information about these relative distances. The model-based
approach explicitly specifies a statistical model of the process that
generated the observed data. Model-based ordination methods (mostly) use
an extension of generalised linear mixed models known as latent variable
models. As the name suggests, these models account for the correlation
between taxa via the inclusion of a small number of latent variables.
Latent variables act as natural ordination axes reflecting unobserved
covariates.

#### [Question:]{style="color:forestgreen"}

#### [32.1 How do the ordinations produced compare to a distance based ordination (for example an NMDS) of the same species data?]{style="color:forestgreen"}

```{r}
#load species abundance data
spe<-as.data.frame(read_xlsx("Data/dune.xlsx", na = ".",sheet = 1, col_names = TRUE)); spe[1]<-NULL
#we need to provide the model with an appropriate probability distribution for the data.
#Given our abundance data is on a scale of 1 - 9 we could classify it as ordinal data,
#but we can try another distribution as well to compare results.

fitnb <- gllvm(y = spe, family = "negative.binomial") #This takes a minute or so to run!
fitord <- gllvm(y = spe, family = "ordinal", zeta.struc = "common")
#ignore the warning messages generated for now!

#As with "normal" linear regressions we can check model fit with widely used tools like
#Akaike Information Criterion (AIC). AIC aims to select the model which best explains 
#the variance in the dependent variable with the fewest number of independent variables
#So it helps select a simpler model over a complex model, if the complexity is not helpful.
#A lower AIC is better!

AIC(fitnb)
AIC(fitord)

#It seems the ordinal model is the best.
```

We can also use graphical checks of model fit, as with any regression.
If you are familiar with these try running plot(fitord) and plot(fitnb).

Now we can look at the ordination plot for the best model.

```{r}
par(mfrow = c(1, 2))
ordiplot.gllvm(
  fitord,
  main = NULL,
  biplot = T,
  s.colors = "black",
  spp.colors = "blue",
  alpha = 0.45,
  cex.spp = 0.7,
  jitter = F,
  s.cex = 0.6
)

abline(h = 0, v = 0, lty = 2)
#compare with an NMDS
distord <- metaMDS(spe)
vegan::ordiplot(distord, type = "t")
par(mfrow = c(1, 1))
```

We can do a lot more with model based ordinations, including
incorporating environmental variables. We can also include information
on traits and perform a fourth corner analysis.

------------------------------------------------------------------------

### [Exercise 33: Fourth Corner]{style="color:cornflowerblue"}

Here we will use another model based method (the "mvabund" package for
analysing multivariate abundance data) to perform a fourth corner
analysis. That is to say we will include trait information and
environmental variables, and investigate the relationship between plant
traits and environmental variables.

#### [Question:]{style="color:forestgreen"}

#### [33.1 Do the correlations in the plot make sense ecologically?]{style="color:forestgreen"}

```{r}
# Read and treat data

env<-as.data.frame(read_xlsx("Data/dune_env.xlsx", na = "NA",sheet = 1, col_names = TRUE)); env[1]<-NULL
env$Management<-as.factor(env$Management)
env$Use<-as.factor(env$Use)
#This is from a database of plant traits
traits<-as.data.frame(read_xlsx("Data/TylerTraits.xlsx", na = "NA",sheet = 1, col_names = TRUE, skip = 1))
rownames(traits)<-traits$Spe; traits$Spe<-NULL 
names(traits)
traits<-traits[,-c(1:6,22, 27,29,30:32)] # Remove trait not feasible for this analysis

#Fourth corner
#Given that there are many traits and even environmental factors that are not strong predictors of abundance, we can reduce the number of #predictor terms in our models.LASSO is an optimization algorithm (Least Absolute Shrinkage and Selection Operator) that is used in this #case.

fitfourth = traitglm(spe, env, traits, method = "glm1path")
fitfourth$fourth #examine fourth corner terms
#and plot the results
a        = max(abs(fitfourth$fourth.corner))
colort   = colorRampPalette(c("blue", "white", "red"))
plot.4th = levelplot(
  t(as.matrix(fitfourth$fourth.corner)),
  xlab = "Environmental Variables",
  ylab = "Species traits",
  col.regions = colort(100),
  at = seq(-a, a, length = 100),
  scales = list(x = list(rot = 45))
)
print(plot.4th)

plot(fitfourth) # for a Dunn-smyth residual plot
qqnorm(residuals(fitfourth)); abline(c(0,1),col="red") # for a normal quantile plot.

#We can test for overall significance using an anova (only on default manyglm model, 
#not available for Lasso model)
fitfourth2 = traitglm(spe, env, traits)
pvalues <- anova.traitglm(fitfourth2,nBoot=10) #overall significant

#We can also explore the effects of individual trait by environment interactions using the 
#summary function ** Warning this is very very slow, even with limited n ** Don't run this 
#last line of code unless you want to take a rather long coffee break!
#We will limit the number of bootstrap reps to 5 (nBoot=5) but 1000 should be used for reliable results/final analyses. Given how slow this is with nBoot=5, I guess this would take many hours...

#summary(fitfourth2, nboot=5)

#Example of output if you don't want to wait for the code to run...
#UsePasture.N_fix1           0.467     0.115    

```

------------------------------------------------------------------------

### [Exercise 34: PLS]{style="color:cornflowerblue"}

PLS regression, like PCA, seeks to find components which maximize the
variability of predictors but differs from PCA as PLS requires the
components to have maximum correlation with the response. The predictor
variables are mapped to a smaller set of variables, and within that
smaller space we perform a regression against the outcome variable. PLS
aims to choose new mapped variables that maximally explain the outcome
variable.

#### [Questions:]{style="color:forestgreen"}

#### [34.1 Which explanatory variables are most important to explain the response variables (A1 and Moisture)?]{style="color:forestgreen"}

#### [34.2 Are there any obvious outliers in the dataset?]{style="color:forestgreen"}

#### [34.3 Does the data seem to be in need of a transformation?]{style="color:forestgreen"}

```{r PLS, echo=TRUE, message=FALSE, warning=FALSE}

library(pls)

#This function takes the same form as a lot of regression models in R:
#plsr(Response variable ~ Explanatory Variables, data = yourdata, scale = TRUE/FALSE)
#However the response variable can be a matrix of variables.

#Here we will use the dune species to predict environmental variables A1 and moisture.
#We have mostly done this in the other direction, but correlations go both ways! 
pls.response <- dplyr::select(dune.env.original, A1, Moisture)
pls.response$Moisture <-  as.numeric(pls.response$Moisture)
pls.response <- as.matrix(pls.response)
pls.exp <- as.matrix(dune)
pls.fit <- plsr(pls.response ~ pls.exp,
                na.action = na.omit,
                validation = "LOO")

#Cross validation is used to help us find the optimal number of retained dimensions.
#Then the model is rebuilt with this optimal number of dimensions.
#Find the number of dimensions with lowest cross validation error
cv <- RMSEP(pls.fit)
best.dims <- which.min(cv$val[estimate = "adjCV", ,]) - 1
best.dims #3 seems optimal

summary(pls.fit)

# Rerun the model with optimal dimensions
pls.fit2 <-
  plsr(pls.response ~ pls.exp, ncomp = best.dims, na.action = na.omit)
#Finally, we extract the useful information and format the output.
coefficients <-  coef(pls.fit2)
sum.coef <-  sum(sapply(coefficients, abs))
coefficients <-  coefficients * 100 / sum.coef
coefficients <-  sort(coefficients[, 1 , 1])
barplot(tail(coefficients, 3))
#The regression coefficients are normalized so their absolute sum is 100 and the result is sorted.

#You can run the following to see the other end of the scale for negative predictors.
barplot(head(coefficients, 3))

summary(pls.fit2)
#pls.fit2$loadings
#We can plot the scores
corrplot(pls.fit2,
         comps = 1:3,
         labels = names(coefficients))

#This gives a pairwise plot of the score values for the three first components.
#Score plots are often used to look for patterns, groups or outliers in the data.
#The scores represents the different locations
plot(pls.fit2, plottype = "scores", comps = 1:3)
#study the predicted vs. measured plot to see if the data needs to be transformed
plot(pls.fit2,
     ncomp = best.dims,
     asp = 1,
     line = TRUE)

#IGNORE THIS CODE # included for backup purposes
#.rs.restartR()
#library(pls)
#library(dplyr)
#library(vegan)
#data("dune.env")
#data("dune")

#pls.response <- dplyr::select(dune.env, A1, Moisture)
#pls.response$Moisture <-  as.numeric(pls.response$Moisture)
#pls.response <- as.matrix(pls.response)
#pls.exp <- as.matrix(dune)
#pls.fit <- plsr(pls.response ~ pls.exp,
#                na.action = na.omit,
#                validation = "LOO")
 

```
