---
title: "MVA Course"
output: 
  html_document:
    df_print: paged
  pdf_document: default
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(eval=FALSE)
library(vegan)
```

Before anything else, install all the packages we will be using. You only need to do this once. Then these can be loaded and used.
```{r echo=TRUE, message=FALSE, warning=FALSE}
###install required packages### 
#delete hash symbol on the following lines and run (only needs to be done first time, not every time you run the code afterwards!)
#install.packages("vegan")
#install.packages("dplyr")
#install.packages("cocorresp")
#install.packages("tibble")
#install.packages("labdsv")
#install.packages("knitr")
#install.packages("dendextend")
#install.packages("pls")
#install.packages("adespatial")
```

### <span style="color:cornflowerblue">First we need to load the R packages we will be using and import the data.</span>

The "dune" dataset is included in the vegan package (an extensive collection of tools for multivariate analysis), but we will be using a slightly modified and extended version of it. In the exercises you may notice that the results returned are sometimes slightly different to those produced by the various graphical interface applications used (Canoco/PAST/Simca). This is a result of different opinions about the best way to implement various design decisions, but the overall patterns seen and the significance of results will be very similar. 

```{r message=FALSE, warning=FALSE}
library(vegan)# Package for multivariate analyses of ecological data
library(dplyr)# Package for data transformation
#Lets's load the data, which is packaged as an RDS object
load("MVM_data.RDS")
#change names to match what is seen in Canoco based presentations 
dune.env <- EnvVar
dune <- Species
#Management factor is character data but we need it to be numeric data for some analyses. 
#For these, we can convert factors to "dummy" variables, that is zeros and ones.
#For other analyses this conversion is done automatically in the background, so we can use the original data. 
#Note:If you are converting a factor that is numeric already you need as.numeric(as.character(x))
#We need to give each management method its own column
dummy_management <- as.data.frame(model.matrix( ~ Management - 1, data=dune.env )) 
#add these to the dataset
dune.env.original <- dune.env #we keep a copy of the original version
dune.env <- dune.env %>% select(A1, Moisture, Manure, Use) %>% cbind(.,dummy_management) 
dune.env$Moisture <- as.numeric(as.character(dune.env$Moisture)) #make numeric
dune.env$Manure <- as.numeric(as.character(dune.env$Manure))
dune.env$Use <- as.numeric(dune.env$Use)
#make column names shorter
dune.env <- dune.env %>% rename(BF = ManagementBF, HF = ManagementHF,
                                NM = ManagementNM, SF = ManagementSF)
```
***
                    

<br>

***

<br>
 
### <span style="color:cornflowerblue">Question 2: Do a PCA on the environmental data related to the Dune meadow dataset.</span>

#### <span style="color:forestgreen">What are the results telling you? In what way do objects/samples to the left differ from objects to the right, and at the bottom from those at the top? Which are the most important gradients in the dataset? Which descriptor variables are related, and which are unrelated? </span>

```{r, warning=FALSE, message=FALSE}
env.pca <- rda(dune.env, scale = TRUE) #vegan uses the same function for PCA and RDA, just depends on if it is constrained or not.
biplot(env.pca)#plot the results
env.pca #summarise results
summary(eigenvals(env.pca)) #see variance explained
```
              

<br>

***

<br>
 

### <span style="color:cornflowerblue">Question 3: For comparison, do also a CA on the Dune Meadow Environmental variables and compare the result with the PCA on the same data! </span>

#### <span style="color:forestgreen">Why do the results in exercise 2 and 3 differ? </span>

```{r, warning=FALSE, message=FALSE}
#unconstrained ordination on environmental data (CA)
env.ca <- cca(dune.env) #vegan uses the same function for CA and CCA
plot(env.ca)
env.ca
summary(eigenvals(env.ca)) #proportion variance explained
```
                    

<br>

***

<br>
 
 


### <span style="color:cornflowerblue">Question 4: Do a CA-ordination on the Dune Meadow species dataset.What are the results telling you? </span>

#### <span style="color:forestgreen">Give a conceptual description on why objects/samples and descriptors/species to the left differ from objects and descriptors to the right, and those at the bottom from those at the top! (Plant ecologists may give a more detailed description, using their knowledge about the species in the dataset). </span>



```{r, warning=FALSE, message=FALSE}
dune.ca <- cca(dune)
plot(dune.ca)
dune.ca
summary(eigenvals(dune.ca)) #proportion variance explained
```
                    

<br>

***

<br>



### <span style="color:cornflowerblue">Question 5: Repeat exercise 4, but with DCA ordination instead. </span>

#### <span style="color:forestgreen">Look at the eigenvalues, the length of gradient, the total variation and the ordination diagram. Explain the differences between results from CA and DCA. </span>

```{r, warning=FALSE, message=FALSE}
dune.dca <- decorana(dune)
dune.dca 
#Detrended correspondence analysis (function decorana).
#Note that we do not get "variation explained" in the R implementation of DCA (and some other functions).
#Here, the developer explains why, "The total amount of variation is undefined in detrended 
#correspondence analysis and therefore proportions from total are unknown and undefined. 
#DCA is not a method for decomposition of variation, and therefore
#these proportions would not make sense either.
plot(dune.dca)
```


<br>

***

<br>
 
### <span style="color:cornflowerblue">Question 6: Add one new species with cover degree 9 at the site with the lowest number of species. Do a new DCA-ordination.</span>

#### <span style="color:forestgreen">What happens with the eigenvalues and ordination diagram?</span>     

```{r ex6, inclue=TRUE, warning=FALSE, message=FALSE}
# calculate the number of species present in each plot
rowSums(Species != 0)
# find the plot with the lowest number of species
which.min(rowSums(Species != 0))
# create a new data frame with a new species with cover degree 9 at site 1
NewSp <- c(9, rep(0,19))
SpeciesNewSp <- data.frame(Species, NewSp)
# run a new DCA with the extra species
DCANewSp <- decorana(SpeciesNewSp)
DCANewSp
```

```{r ex6plot, inclue=TRUE, warning=FALSE, message=FALSE}
# plot the DCA
ordiplot(DCANewSp, type="text", main="DCA with new sp.")
```


You may be surprised that you haven't got any total inertia values when printing decorana results, 
although in other software (e.g. CANOCO) these are available, together with percentage variance 
explained by particular axes. The reason for this is that DCA does not support the concept of total 
inertia values (also, it produces only four axes, i.e. four eigenvalues). See the comment in the 
code for the question 5.
                    

<br>

***

<br>
 

### <span style="color:cornflowerblue">Question 7:  Repeat the ordination in exercise 6, but make the new species supplementary.</span>

#### <span style="color:forestgreen"> Compare eigenvalues and ordination diagrams between exercises 5, 6 and 7! Explain the differences! What is the effect of making a species supplementary?</span>

```{r ex7, inclue=TRUE, warning=FALSE, message=FALSE}
```

Skip this question as supplementary species are rarely used and this method is not available for DCA in R.




<br>

***

<br>

### <span style="color:cornflowerblue">Question 8:	Add one new sample to the data, with the existing species Chenopodium album "Che_alb" having a cover of 9. Make a new DCA-ordination. </span>

#### <span style="color:forestgreen"> What happens with the eigenvalues and ordination diagram? Why did we choose to add Chenopodium album in the new sample? </span>

```{r ex8, inclue=TRUE, warning=FALSE, message=FALSE}
# figure out the column number for Chenopodium album (abbreviated "Chenalbu")
which(colnames(dune)=="Che_alb")
# create a new data frame with a new sample containing Chenopodium album with a cover of 9
NewSample <- c(rep(0,7), 9, rep(0,22))
SpeciesNewSample <- data.frame(rbind(dune, NewSample))
# run a new DCA with the extra sample
DCANewSample <- decorana(SpeciesNewSample)
DCANewSample
# plot the DCA
ordiplot(DCANewSample, type="text", main="DCA with new sample Chenopodium album")
```



<br>

***

<br>

### <span style="color:cornflowerblue">Question 9:	Keep the new sample but delete the added Che_alb=9, and instead add Leo_aut = 5 to the new sample. Do a new DCA ordination!  </span>


#### <span style="color:forestgreen"> What happens with the eigenvalues? Look at the ordination diagram. Where is the extra sample located? </span>

```{r ex9, inclue=TRUE, warning=FALSE, message=FALSE}
# figure out the column number for Chenopodium album (abbreviated "Chenalbu")
which(colnames(dune)=="Leo_aut")
# create a new data frame with a new sample containing Chenopodium album with a cover of 9
NewSample <- c(rep(0,15), 5, rep(0,10))
SpeciesNewSample <- data.frame(rbind(dune, NewSample))
# run a new DCA with the extra sample
DCANewSample <- decorana(SpeciesNewSample)
DCANewSample
# plot the DCA
ordiplot(DCANewSample, type="text", main="DCA with new sample Leontodon autumnalis")
```


<br>

***

<br>

### <span style="color:cornflowerblue">Question 10:	Downweighting of rare species. Do a new DCA on the original species data table, but with downweighting of rare species. </span>

#### <span style="color:forestgreen"> What is the effect of downweighting? Compare eigenvalues and ordination diagrams from Exercise 5 and 10, and explain the differences! </span>

```{r ex10, inclue=TRUE, message=FALSE, warning=FALSE}
# DCA on original data set but with argument iweigh	(0: no)
DCADownweight <- decorana(dune, iweigh=1)
DCADownweight

## original DCA
dune.dca

# plot the DCAs
ordiplot(DCADownweight, type="text", main="DCA with downweighted rare species")

ordiplot(dune.dca, type="text", main="DCA original")
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 11:	Non Metric Multidimensional Scaling </span>

#### <span style="color:forestgreen"> Compare the NMDS ordination with the ordinations obtained using CA and DCA! What is the stress value for the NMDS? </span>

```{r ex11, inclue=TRUE, warning=FALSE, message=FALSE}
# NMDS on original data set with 2 reduced dimensions
NMDS <- metaMDS(dune, k=2)
NMDS
# plot the NMDS
ordiplot(NMDS, type="text", main="NMDS")

#plot the CA
ordiplot(dune.ca, type="text", main="DCA")

#plot the DCA
ordiplot(dune.dca, type="text", main="DCA")

```

<br>

***

<br>


### <span style="color:cornflowerblue">Question 12:	Procustes rotation and Co-correspondence analysis  </span>

#### <span style="color:forestgreen"> Are the two data sets (species and bugs) correlated? Give a very short ecological interpretation of the results! </span>

These are two ways of comparing ordinations or the structure in related multivariate datasets. They are used to compare how objects differ when they are described by different sets of descriptors, or if there are repeated samplings of the same objects using the same descriptors.

```{r ex12, inclue=TRUE, warning=FALSE, message=FALSE}
library(cocorresp)
# DCA on the plant species data 
DCASpecies <- decorana(dune)
# DCA on the insect data (called Bugs)
DCABugs <- decorana(Bugs)
DCABugs
# procrustes analysis on the two DCAs from different datasets
ProC <- procrustes(DCASpecies, DCABugs)

plot(ProC, kind=1, type="text")
# Labels show the position of the samples in the second ordination, and arrows point to their positions in the target ordination
# Function protest tests the non-randomness (`significance') between two configurations
ProCsig <- protest(dune.dca, DCABugs)
ProCsig
# Co-correspondence analysis
CoCor <- coca(dune ~ ., data = Bugs, method = "symmetric")
summary(CoCor)
corAxis(CoCor)
```


<br>

***

<br>

## Exercises direct methods

### <span style="color:cornflowerblue">Question 13:	Do an RDA and a CCA on the Dune Meadow data set, using the full set of environmental variables.  </span>

#### <span style="color:forestgreen"> Compare the ordination diagrams and numerical results! What does the significance test tell you? When only looking at the results, which method explains the variation best? Taking the nature of the data into account, which method should be used?  </span>

```{r ex13, inclue=TRUE, warning=FALSE, message=FALSE}
# RDA 
dune.rda <- rda(dune ~ . , dune.env.original)
summary(dune.rda)
anova(dune.rda)

# CCA 
dune.cca <- cca(dune ~ . , dune.env.original)
summary(dune.cca)
anova(dune.cca)

#plots
ordiplot(dune.rda, type="text", main="RDA")
ordiplot(dune.cca, type="text", main="CCA")
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 14:	Use the results from the CCA ordination in exercise 13 to look at and interpret the variance inflation factor, and to look for warnings about outliers in the explanatory data!   </span>

#### <span style="color:forestgreen"> Look at and interpret the inflation factors for each environmental variable! Are there any extreme values? </span>

```{r ex14, inclue=TRUE, warning=FALSE, message=FALSE}
# VIF for CCA 
dune.cca <- cca(dune ~ . , dune.env)
vif.cca <- vif.cca(dune.cca)
vif.cca
```
Function vif.cca gives the variance inflation factors for each constraint or contrast in factor constraints. In partial ordination, conditioning variables are analysed together with constraints. Variance inflation is a diagnostic tool to identify useless constraints. A common rule is that values over 10 indicate redundant constraints. 

<br>

***

<br>

### <span style="color:cornflowerblue">Question 15:	Do a CA on the Dune Meadow data set, with the full set of environmental variables included in the analysis, and compare with a CCA on the same data sets. </span>

#### <span style="color:forestgreen"> Compare the constrained and unconstrained ordinations! What is the difference between the methods? When could an unconstrained ordination with explanatory variables be a good option? </span>


```{r ex15, inclue=TRUE, warning=FALSE, message=FALSE}
# CA with env. variables 
fit <- envfit(dune.ca, dune.env, perm=999)
fit
ordiplot(dune.ca,  type="text", main="CA with env. variables fitted after")
plot(fit)

#CCA 
dune.cca <- cca(dune ~ . , dune.env.original)
dune.cca
ordiplot(dune.cca, type="text", main="CA")
```

<br>

***

<br>

### <span style="color:cornflowerblue"> Question 16:	Next, you should investigate whether the four different management regimes (SF standard farming, BF biodynamical farming, HF hobby farming, and NM nature management) have any effect on the species distribution. This is done by using these four management types as environmental variables in a constrained ordination.  </span>

#### <span style="color:forestgreen"> Inspect and interpret the results of the permutations and graph! Compare with the CCA on the full set of explanatory data in exercise 13. </span>

To investigate if the vegetational pattern is an effect of management, or if the pattern is random, Monte Carlo permutations will be used. 


```{r ex16, inclue=TRUE, warning=FALSE, message=FALSE}
# CCA using only Management as the environmental variable
dune.ccaMN <- cca(dune ~ Management , dune.env.original)
summary(dune.ccaMN)
# Permutation test on CCA
MCperm <- permutest(dune.ccaMN, permutations=999)
MCperm
# Plot the ordination with only Management
ordiplot(dune.ccaMN, type = "points", main = "CCA only Management")
ordihull(dune.ccaMN,groups = dune.env.original$Management,draw = "polygon",col = "grey70", label = T)
```

<br>

***

<br>


### <span style="color:cornflowerblue">Question 17:	Next, you want to investigate whether initial soil characteristics (A1 and moisture) also are important for differences in species composition between fields? To do this, do another CCA with A1 and Moisture as environmental variables and perform a significance test.   </span>

#### <span style="color:forestgreen"> Explain the results of the permutations! Compare the results in exercises 16 and 17! </span>

```{r ex17, inclue=TRUE, warning=FALSE, message=FALSE}
# CCA using A1 and moisture as the environmental variable
dune.ccaA1ms <- cca(dune, dune.env[,1:2])
summary(dune.ccaA1ms)
# Permutation test on CCA
MCpermA1ms <- permutest(dune.ccaA1ms, permutations=999)
MCpermA1ms 
# Plot the ordination
ordiplot(dune.ccaA1ms, type="text", main="CCA A1 and moisture")

```

<br>

***

<br>

## Direct methods with testing of environmental variables

### <span style="color:cornflowerblue"> Question 18: Partial ordination    </span>

Now, you have concluded that both soil characteristics and management type are determining the species composition significantly. But, what if the observed differences between management types are not caused by management type but by initial differences in soil characteristics? To investigate this, you have to test whether there still is a difference in vegetation between management types, after accounting for (i.e. removing) the effects of soil characteristics.
This is done by a partial CCA ordination. Partial ordinations are used to eliminate effects of selected variables by specifying them as covariates. In this case, use the two soil characterising variables (A1 and Moisture) as covariates, and the four management types as environmental variables. 


#### <span style="color:forestgreen">What does the permutation test tell you? Also, compare the results with the results from the permutation test on only management types!  </span>

```{r ex18, inclue=TRUE, warning=FALSE, message=FALSE}
#This partials out the effect of A1 and Moisture before analysing the effects of management
MN.cca <- cca(dune ~ Management + Condition(A1 + Moisture), data=dune.env.original)
summary(MN.cca)
#permutation test
permutest(MN.cca, permutations=999)
# Plot the ordination
ordiplot(MN.cca, type = "points")
ordihull(MN.cca,groups = dune.env.original$Management,draw = "polygon",col = "grey70", label = T)
```

<br>

***

<br>

### <span style="color:cornflowerblue"> Question 19: The amount of applied manure is a third important variable given in the environmental data set associated with the dune meadow data set. Repeat exercise 18, but with manure added to the set of covariates. </span>


#### <span style="color:forestgreen"> Inspect and interpret the results of the permutations! </span>

```{r ex19, inclue=TRUE, warning=FALSE, message=FALSE}
#This partials out the effect of A1, Moisture, and Manure before analysing the effects of management
MN.cca2 <- cca(dune ~ Management + Condition(A1 + Moisture + Manure), data=dune.env.original)
summary(MN.cca2)
#permutation test
permutest(MN.cca2, permutations=999)
```

<br>

***

<br>

### <span style="color:cornflowerblue"> Question 20: Forward selection </span>

#### <span style="color:forestgreen"> We have now concluded that management type is of subordinate importance. Now, you want to test the importance or significance of individual environmental variables. This is done by Forward selection. What is the effect of Forward selection? In what way do these results differ from the results obtained when the whole Environmental data table was used as explanatory variables? Were the set of explanatory variables selected by the Forward selection the best set? Why would one perform a p-value correction?  </span>

```{r ex20, echo=TRUE, message=FALSE, warning=FALSE, inclue=TRUE}
mod0 <- rda(dune ~ 1, dune.env)  # Model with intercept only
mod1 <- rda(dune ~ ., dune.env)  # Model with all explanatory variables
fs <- ordistep(mod0, scope = formula(mod1), direction="forward")
```
Since there is (a potentially high) number of tests of significance during the forward selection procedure, 
it is better to apply a correction for multiple testing issue. Here we apply the Holm correction.
```{r}
n.tests <- ncol(dune.env)  # number of tests equals number of variables which are being selected
pval.adj <- p.adjust(fs$anova$`Pr(>F)`, method = 'holm', n = n.tests)
fs$anova$`Pr(>F)` <- pval.adj
fs$anova
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 21: Interpreting ordination results using species traits </span>

#### <span style="color:forestgreen"> Experienced plant ecologists may already have looked at the species in the Dune Meadow ordination graphs and concluded that species with similar traits occur together. This is possible if you have good knowledge about plant species ecological preferences, and if there are relatively few species in your dataset. In this exercise we will use tabulated data on species ecological preferences (the Ellenberg indicator values) to interpret the results of the ordinations. The Ellenberg indicator values are described on the first page in this booklet. Which Ellenberg values are most important for the distribution of the species? What do the different axes represent in terms of environmental gradients? </span>

```{r, warning=FALSE, message=FALSE}
#rename data with Ellenberg values to match Canoco based presentations
dune.ell <- Ellenberg
dune.mean.ell <- Mean_Ellenberg
#The vegan implementation of forward selection of variables does not accept NA values.
#To get around this we will replace the two NAs with the mean value for all plots in that column.
#Do not do this in a real analysis without seriously thinking about how it affects your results!
#This also means that the ordinations you produce may look slightly different to those produced in Canoco, 
#but the overall patterns and significance of results will be the same.
## Code to replace missing values with the column mean, and save the results as a new dataframe
## called dune.mean.ell.impute (since we are imputing the missing values).
dune.mean.ell.impute <- data.frame(
    sapply(
        dune.mean.ell,
        function(x) ifelse(is.na(x),
            mean(x, na.rm = TRUE),
            x)))
#CCA analysis
#create global model with CCA (including all variables) and test it's significance, and if it is significant,
#we use the ordistep function with appropriate arguments to do forward selection of variables.
cca1 <- cca(dune ~ ., data = dune.mean.ell.impute) # full model (with all explanatory variables)
anova(cca1) #overall model is significant
# Start by making ordinations
cca0 <- cca (dune~ 1, data = dune.mean.ell.impute) # empty model only with intercept
cca1 <- cca(dune ~ ., data = dune.mean.ell.impute, na.action = na.omit) # full model (with all explanatory variables)
cca1
plot(cca1, main="CCA, all data")
#Stepwise approach, using "ordistep"
step2<-ordistep(cca0, scope = formula(cca1), direction="forward")
step2$anova # which three are most significant in the final selection?
```

#### 

<br>

***

<br>



### <span style="color:cornflowerblue">Question 22: Decomposition of variance </span>

#### <span style="color:forestgreen"> Perform the analyses and interpret the results! How much of the total variation (% of All) is explained by: - uniquely by Management (effect of Soil removed), - uniquely by Soil  (effect of Management removed), - jointly by Soil and Management (the interaction) - Not by Soil or Management? </span>

In usual analysis of variance experiments, the variance is decomposed into components. The same can be done in multivariate analyses. In this exercise you will decompose the variance in the Dune meadow data set into different variance components. You will use two groups of variance components: Group 1 is Management and Group 2 is the soil variables (A1 and Moisture). In ordinations, the variance is expressed by the sum of the eigenvalues. 

The result gives you the fraction of the total variation that is explained by:
    a) uniquely by Management (effect of soil removed),
    b) uniquely by soil (effect of Management removed),
    c) jointly by Soil and Management (the interaction).
    
The total explained variation is the sum of a), b) and c). 

The total variation in the dataset is the sum of all unconstrained eigenvalues.


```{r, warning=FALSE, message=FALSE}
## variance partitioning
management <- dune.env[,c("BF", "HF", "NM", "SF")]
soil <- dune.env[,c("A1", "Moisture")]
# examine the explanatory variable of each class of variables.
varp <- varpart(dune, management, soil)
varp
plot (varp, digits = 2, Xnames = c('Management', 'Soil'), bg = c('red', 'blue'))
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 23: Diagnostic species for a priori defined groups </span>

#### <span style="color:forestgreen"> Use the resulting data table to find the two top indicator species for each of the four Management types! </span>

In this exercise we will investigate if any particular species are indicative for the four different management types. The CANOCO version of this analysis is a simplified version of these types of tests. The full versions also give significance testing and more comprehensive result presentations. See for instance IndVal by Dufresne & Legendre.
    1. Go to menu Data /Add new table(s) / Indicator values.
    2. Mark the Species data table, and Management as the factor defining groups.
    3. Opt for Indicator Value (quantitative). 
    4. Export the table to Excel to be able to sort the data to find the highest values = best indicators for a group.

```{r, warning=FALSE, message=FALSE}
library(labdsv)
library(tibble)
iva <- indval(dune, dune.env.original$Management) #this time we don't need dummy variables for management
iva.df <- as.data.frame(iva$indval)
#arrange in descending order to find best indicator species, change BF to other management types as needed
arrange(rownames_to_column(iva.df), desc(HF))
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 24: Analysing a time series with vegetation data </span>

#### <span style="color:forestgreen"> Is there a significant change in species composition over time?  What is the effect of defining Plots as cofactors? </span>

Quite often vegetation ecologists have data from repeated inventories in permanent plots. The overall question to answer is if there has been a significant and directional change in the species composition. In this exercise we have rearranged the Dune Meadow data so that it consists of only 10 plots, each analysed 2 times (same species as in all other exercises, just grouping and rearrangement of plots). We want to analyse if there has been a consistent change in species composition between the two inventories. We are thus not interested in differences between the ten plots. 

```{r, warning=FALSE, message=FALSE}
#This partials out the effect of Plot before analysing the effects of Time
time.cca <- cca(SpeTS ~ Time + Condition(Plot), data=EnvTS)
time.cca
treat <- EnvTS$Time #
colvec <- c("red2", "green4") #set colours to be applied to different levels of factor "Time"
plot(time.cca, type = "n", display = c("lc"))#plots the axes
with(time.cca, points(time.cca, display = c("lc"), col = colvec[treat],#plots the points
                      pch = 21, xlim = c(-2,2), bg = colvec[treat]))
```

```{r}
#permutation test
with(EnvTS, anova(time.cca, by="term", perm=500, strata=Plot))
```


```{r, warning=FALSE, message=FALSE}
#Compare with results for time as only explanatory variable and no cofactors
time2.cca <- cca(SpeTS ~ Time, data=EnvTS)
time2.cca
#permutation test
with(EnvTS, anova(time2.cca, by="term", perm=500))
```


```{r, warning=FALSE, message=FALSE}
## plot ellipsoid hulls
treat <- EnvTS$Time
plot(time2.cca, type = "n")#plots the axes
with(time2.cca, points(time2.cca, col = colvec[treat],#plots the points
                      pch = 21, bg = colvec[treat]))
ordihull(time2.cca,groups=treat,draw="polygon",col="grey70",label=T)#draw hulls around area of each level of factor "Time"
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 25: A multivariate Before-After-Control- Impact (BACI) study </span>

#### <span style="color:forestgreen"> Did the treatment have an effect? What was permuted? </span>


In this exercise we will continue to use the rearranged Dune Meadow data from exercise 24. A difference is that the plots are now divided into four groups:
    1. Control plots before a treatment (i.e. not treated)
    2. Control plots after a treatment (i.e. not treated)
    3. Impact plots before treatment (i.e. not treated)
    4. Impact plots after treatment (this is where the treatment is made)

The four groups are indicated in variable Treat in the environmental data set (EnvTS).
The question you want to answer is if the treatment caused a change in species composition that is significantly different from the change in the control plots.


```{r, warning=FALSE, message=FALSE}
baci.cca <- cca(SpeTS ~ Treat + Condition(Plot + Time), data=EnvTS)
baci.cca
```



```{r, warning=FALSE, message=FALSE}
with(EnvTS, anova(baci.cca, by="term", perm=500, strata=Treat))
#Here with() is a special function that makes variables in dune.env visible to
#the following command. If you only type Moisture in an R prompt, you will get
#an error of missing variables
```




```{r, warning=FALSE, message=FALSE}
## plot results
treat <- EnvTS$Treat
ordiplot(baci.cca,type="points")
ordihull(baci.cca,groups=treat,draw="polygon",col="grey70",label=T)#draw hulls around area of each level of factor "Time"
```


<br>

***

<br>



### <span style="color:cornflowerblue">Question 26 : Classification </span>

#### <span style="color:forestgreen"> How do the different clustering techniques differ? </span>


Start by testing the non-hierarchical K-means clustering, using Multivar / K-Means. Note that it is recommended that K-means need at least 100 observations (500 according to some sources) to be reliable! Let us ignore the sample size issue for the moment, and ask for 4 clusters (for later comparison with the four management types). Test different hierarchical agglomeration algorithms and similarity indices. Use at least the Euclidian and the Bray-Curtis similarity measures for the hierarchical clustering technique. Try also the Ward's method! 

```{r, warning=FALSE, message=FALSE}
library(dendextend)
#make a distance matrix using the Bray-Curtis method
dune.dist <- vegdist(dune, method = "bray")
# K-Means Cluster Analysis
fit <- kmeans(dune, 4, nstart = 50) # we specify a 4 cluster solution
#As the final result of k-means clustering result is sensitive to the random starting assignments, we specify nstart = 50. This means #that R will try 50 different random starting assignments and then select the best results corresponding to the one with the lowest #within cluster variation.
#Note that k-means clustering involves randomness so you won't neccessarily get exactly the same results if you repeat it.
#There are also several different algorithms that can be used in k-means clustering, and different software 
#have can use different defaults.Check out the help file for the kmeans function if you want to know more about these.
#cluster number for each data point
fit$cluster
#cluster sizes
fit$size
#compare clusters to management category
table(fit$cluster, dune.env.original$Management)
#try some approaches to hierarchical clustering
dend2 <- dune %>% # data
        dist(method = "euclidean") %>% # calculate a distance matrix, choose method 
        hclust(method = "ward.D") %>% # Hierarchical clustering, choose method 
        as.dendrogram # Turn the object into a dendrogram.
dend2 %>% set("labels_color") %>% set("branches_lwd") %>%  set("branches_k_color", k = 4) %>% plot
#the dist() function does not include Bray-Curtis but we have already made a distance matrix using 
#the vegdist() function from vegan.
dend3 <- dune.dist %>% # data (Bray-Curtis)
        hclust(method = "aver") %>% # Hierarchical clustering, choose method 
        as.dendrogram # Turn the object into a dendrogram.
dend3 %>% set("labels_color") %>% set("branches_lwd") %>%  set("branches_k_color", k = 4) %>% plot
#we can colour the labels based on their grouping in the kmeans analysis to compare clusters
dend3 %>% set("labels_color", fit$cluster) %>% set("branches_lwd") %>%  set("branches_k_color", k = 4) %>% plot
#tanglegram is a nice function to compare dendrograms that you might want to try out as an extra
tanglegram(dend2, dend3)
```

<br>

***

<br>

### <span style="color:cornflowerblue">Question 27: ANOSIM </span>

#### <span style="color:forestgreen"> What does the result tell you? What other types of predefined clusters could you use? </span>

ANOSIM (ANalysis Of Similarities) is a non-parametric test of significant difference between two or more groups, based on any distance measure. In this case, use the clusters from the K-means exercise. Change to Bray-Curtis similarity index.



```{r, warning=FALSE, message=FALSE}
dune.dist <- vegdist(dune, method = "bray") #create distance matrix based on Bray-Curtis method
dune.ano <- anosim(dune.dist, fit$cluster) #compare groupings based on cluster analysis
summary(dune.ano)
#plot(dune.ano)
#title(sub = "Anosim, groups = cluster analysis")
#ANOSIM gives you the P value and a R value. R value close to 1 indicates high separation between levels of your factor while R value #close to 0 indicate no separation between levels of your factor.
```

```{r warning=FALSE}
dune.ano2 <- anosim(dune.dist, dune.env.original$Management) #try comparing groupings based on management instead
summary(dune.ano2)
#plot(dune.ano2, sub = "Anosim, groups = management")
```


<br>

***

<br>


### <span style="color:cornflowerblue">Question 28: SIMPER </span>

#### <span style="color:forestgreen"> Compare the different groups and check which species that contributes mostly to the difference between pairs of clusters.  </span>

After a significant ANOSIM, you may want to know which species are primarily responsible for the observed difference between clusters. SIMPER (Similarity Percentage) will do this for you. The test does not come with significance testing. In the output tables, taxa are sorted in descending order of contribution to group difference. 
```{r, warning=FALSE, message=FALSE}
sim <- simper(dune, fit$cluster) #start with kmeans clusters
summary(sim)
```
```{r}
sim2 <- simper(dune, dune.env.original$Management) #try management groupings
summary(sim2)
```


<br>

***

<br>

### <span style="color:cornflowerblue">Question 29: PLS </span>

#### <span style="color:forestgreen"> Examine the p-values to explore which response variables actually give a significant result. </span>

Unless you are already familiar with PLS in R we suggest that you follow the exercise in SIMCA and then try performing the same analysis by modifying the R code below. Remember you can always read the helpfile for any R function by entering ? and the function name in the console, such as ?plsr in this case.

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
data.pls <- Trend_Lakes_2015_PLS
#View(data.pls)
library(pls)
#PLS regression, like PCA, seeks to find components which maximize the variability of predictors but differs from PCA as PLS requires #the components to have maximum correlation with the response.
#Change 'Antal arter' (number of species in Swedish) to your choice of response variable(s)
pls.fit <- plsr(`Antal arter` ~ ., data = data.pls[c(9:35)], scale = TRUE, validation = "CV")
#The predictor variables are mapped to a smaller set of variables, and within that smaller space we perform a regression against the #outcome variable. PLS aims to choose new mapped variables that maximally explain the outcome variable.
#The next step is to remove unwanted variables and then build a model.  
#Cross validation is used to find the optimal number of retained dimensions.
#Then the model is rebuilt with this optimal number of dimensions. 
#Find the number of dimensions with lowest cross validation error
cv <- RMSEP(pls.fit)
best.dims <- which.min(cv$val[estimate = "adjCV", , ]) - 1
best.dims 
# Rerun the model
pls.fit2 <-  plsr(`Antal arter` ~ ., data = data.pls[c(6:35)], ncomp = best.dims)
#Finally, we extract the useful information and format the output.
coefficients <-  coef(pls.fit2)
sum.coef <-  sum(sapply(coefficients, abs))
coefficients <-  coefficients * 100 / sum.coef
coefficients <-  sort(coefficients[, 1 , 1])
barplot(tail(coefficients, 5))
#The regression coefficients are normalized so their absolute sum is 100 and the result is sorted.
#The results below show that 'Tot-N_TNb' and 'Mn' are the two most importnat positive predictors of 'Antal arter'.  You could run the code #barplot(head(coefficients, 5)) to see that at the other end of the scale for negative predictors.
summary(pls.fit2)
plot(pls.fit)
validationplot(pls.fit2, val.type = "MSEP")
predplot(pls.fit2)
coefplot(pls.fit2)
```